Formalized Linear Programming

### 2019-09-27 TeX/paper.tex
Minor grammar corrections

In the tableau representation of this problem, the vectors $x_r$ and $x_c$ are viewed as minors of a single variables $x$. The subset of $\mathbb{Q}^{m+n}$, $\{x : x_c = Ax_r + b \wedge x_c \ge 0 \wedge x_r \ge 0\}$ is stored as part of the simplex algorithm. The tableau consists of a matrix $A$ of coefficients, a column vector $b$, a reduced cost vector $c$, and the constant part of the objective function $z$. Additionally two vectors of indices $r$ and $c$ defining the minors $x_r$ and $x_c$ are also stored.
\begin{equation}
  \begin{aligned}
    x_{r_1} && = && A_{11}x_{c_1} && + && A_{12}x_{c_2} && + && \dots && + && A_{1n}x_{c_n} && + && b_1 \\
    x_{r_2} && = && A_{21}x_{c_1} && + && A_{22}x_{c_2} && + && \dots && + && A_{2n}x_{c_n} && + && b_2 \\
    \vdots && && \vdots && && \vdots && && && && \vdots && \vdots \\
    x_{r_m} && = && A_{m1}x_{c_1} && + && A_{m2}x_{c_2} && + && \dots && + && A_{mn}x_{c_n} && + && b_n \\
    c^Tx_c + z && = && c_1x_{c_1} && + && c_2x_{c_2} && + && \dots && + && c_nx_{c_n} && + && z
  \end{aligned}
\end{equation}
In this tableau representation the $x_r$ are known as the \textit{row} variables, and the $x_c$ are the \textit{column} variables. In a tableau the row variables are expressed as affine functions in terms of the column variables.

@@

The choice of pivot row and column satisfying (\ref{eqn:valid_pivot}) is non-unique, and must be chosen carefully in order to guarantee termination. Different implementations of the simplex algorithm use different pivot rules to guarantee termination. If $z$ was strictly increasing it would straightforward to prove the algorithm terminates, since there are only finitely many partitions of the variables into row and column variables, and the simplex algorithm cannot repeat a partition if $z$ is strictly increasing. However when $b_i = 0$ in the pivot row $i$, $z$ does not increase, and ensuring termination is more difficult. Our implementation uses Bland's Rule to ensure termination. \cite{Blands_rule}.

@@

The simplex algorithm was originally written on an extremely slow implementation of matrices. The implementation in the Lean maths library \color{red}[cite mathlib]\color{black} is as the type of binary functions from two finite sets into a ring, in our case $\mathbb{Q}$. This implementation is convenient for writing proofs, but very slow to compute. In order to have a practical executable version of the simplex algorithm, it was necessary to transfer the algorithm and proofs onto a faster implementation of matrices using arrays. Ideally this would be done using a refinement framework \color{red} brief explanation of what that means here. Would it be possible to generalize proof onto a type class rather than a concrete type using this framework?\color{black}. This is not currently available in Lean, so it was instead done in an ad hoc way, using a tableau type class. The proof of the simplex algorithm can be generalized onto any type for which there is an instance of the \lstinline|is_tableau| type class.

>>>>

In the tableau representation of this problem, the vectors $x_r$ and $x_c$ are viewed as minors of a single variable $x$. The subset of $\mathbb{Q}^{m+n}$, $\{x : x_c = Ax_r + b \wedge x_c \ge 0 \wedge x_r \ge 0\}$ is stored as part of the simplex algorithm. The tableau consists of a matrix $A$ of coefficients, a column vector $b$, a reduced cost vector $c$, and the constant part of the objective function $z$. Additionally two vectors of indices $r$ and $c$ defining the minors $x_r$ and $x_c$ are also stored.
\begin{equation}
  \begin{aligned}
    x_{r_1} && = && A_{11}x_{c_1} && + && A_{12}x_{c_2} && + && \dots && + && A_{1n}x_{c_n} && + && b_1 \\
    x_{r_2} && = && A_{21}x_{c_1} && + && A_{22}x_{c_2} && + && \dots && + && A_{2n}x_{c_n} && + && b_2 \\
    \vdots && && \vdots && && \vdots && && && && \vdots && \vdots \\
    x_{r_m} && = && A_{m1}x_{c_1} && + && A_{m2}x_{c_2} && + && \dots && + && A_{mn}x_{c_n} && + && b_n \\
    c^Tx_c + z && = && c_1x_{c_1} && + && c_2x_{c_2} && + && \dots && + && c_nx_{c_n} && + && z
  \end{aligned}
\end{equation}
In this tableau representation the $x_r$ are known as the \textit{row} variables, and the $x_c$ are the \textit{column} variables. In a tableau the row variables are expressed as affine functions in terms of the column variables.

@@

The choice of pivot row and column satisfying (\ref{eqn:valid_pivot}) is non-unique, and must be taken carefully in order to guarantee termination. Different implementations of the simplex algorithm use different pivot rules to guarantee termination. If $z$ was strictly increasing it would be straightforward to prove the algorithm terminates, since there are only finitely many partitions of the variables into row and column variables, and the simplex algorithm cannot repeat a partition if $z$ is strictly increasing. However when $b_i = 0$ in the pivot row $i$, $z$ does not increase, and ensuring termination is more difficult. Our implementation uses Bland's Rule~\cite{Blands_rule} to ensure termination.

@@

The simplex algorithm was originally written on an extremely slow implementation of matrices. The implementation in the Lean maths library \color{red}[cite mathlib]\color{black} is as the type of binary functions from two finite sets into a ring, in our case $\mathbb{Q}$. This implementation is convenient for writing proofs, but very slow to compute. In order to have a practical executable version of the simplex algorithm, it was necessary to transfer the algorithm and proofs onto a faster implementation of matrices using arrays. Ideally this would be done using a refinement framework \color{red} brief explanation of what that means here. Would it be possible to generalize this proof onto a type class rather than a concrete type using this framework?\color{black}. This is not currently available in Lean, so it was instead done in an ad hoc way, using a tableau type class. The proof of the simplex algorithm can be generalized onto any type for which there is an instance of the \lstinline|is_tableau| type class.
